Analyses of 27 years of stand structure development and basal area growth response to a range of initial basal area density levels, 1992-2019

### 2020-05-03 manuscript/EP1162_manuscript_draft.Rmd
more manny work

@@

@@

# Create figures of raw data to put into Appendix
pi.Init.rawPlot<-  
  piDat %>% 
  ggplot(aes(x=log(Init),y=log(PI),color=Species))+
  geom_point()+
  geom_smooth(method="lm",se=FALSE,color="black")+
  facet_grid(Species~BA.Target)  +
   ylab("log(BA increase, m²)")+
  xlab("log(Tree BA in 1992, m²)")+
    theme_bw()+
  theme(legend.position = "none")

piPCT.Init.rawPlot<-  
  piDat %>% 
  ggplot(aes(x=log(Init),y=log(PIpct),color=Species))+
  geom_point()+
  geom_smooth(method="lm",se=FALSE,color="black")+
  facet_grid(Species~BA.Target) +
  ylab("log(% BA increase)")+
  xlab("log(Tree BA in 1992, m²)")+
  theme_bw()+
  theme(legend.position="none")

@@

# Create figure showing marginal means from model
fitBApct.plot<-
  ggpredict(fitBApct,terms=c("Init [exp]","Species","BA.Target")) %>% 
  dplyr::rename(Species=group) %>% 
  ggplot(aes(x,predicted,color=Species))+
  geom_line()+
  facet_wrap(~facet)+
  ylab("Predicted 27-year basal area % increase")+
  xlab("Tree BA in 1992 (m²)")+
  theme_bw()+
  theme(legend.position="bottom")

@@

# Raw plots
baAnn.raw.plot<-
  # Plot boxplot of increment by year
  piDat %>% 
  ggplot(aes(x=Year,y=Inc,color=BA.Target))+
  geom_boxplot()+
  facet_grid(Species~BA.Target)+
  xlab("Measurement year")+
  ylab("Annualized basal area increment (m²)")+
  theme(legend.position = "none")+
    theme_bw()+
  theme(legend.position = "none")

@@

@@

`r figNums("fitBA.Ann.plot",caption=paste("Predicted annualized basal area increment (m^2^ at breast height) over four periods from a linear mixed effects model explaining ", MuMIn::r.squaredGLMM(fitBA.Ann)[1] %>% round(3)*100, "% variation in log-transformed basal area increment (not shown) with period, treatment unit and species as fixed effects (p<0.001 for all fixed effects, now shown). Basal area increment are average for growing seasons within each period.  For example, values at x-axis value of 1994 are mean for three growing seasons (between 1992 and 1994).  Predictions are back-transformed to original response scale.  Error bars represent 95% confidence intervals.",sep=""))`

@@

>>>>

# Set ggplot theme
theme_set(theme_bw())

@@

#### Volume estimates
Tree volumes were estimated using species- and region-specific equations [@Nigh2016] based on tree height and DBH.  We estimated tree heights by fitting diameter-height models developed separately for fir and spruce, and for each period.  Trees with broken tops were omitted from diameter-height models.

@@

# Create figures of raw data to put into Appendix
pi.Init.rawPlot<-  
  piDat %>% 
  ggplot(aes(x=log(Init),y=log(PI),color=Species))+
  geom_point()+
  geom_smooth(method="lm",se=FALSE,color="black")+
  facet_grid(Species~BA.Target)  +
   ylab("log(BA increase, m²)")+
  xlab("log(Tree BA in 1992, m²)")+
  theme(legend.position = "none")

piPCT.Init.rawPlot<-  
  piDat %>% 
  ggplot(aes(x=log(Init),y=log(PIpct),color=Species))+
  geom_point()+
  geom_smooth(method="lm",se=FALSE,color="black")+
  facet_grid(Species~BA.Target) +
  ylab("log(% BA increase)")+
  xlab("log(Tree BA in 1992, m²)")+
  theme(legend.position="none")

@@

# Create figure showing marginal means from model
fitBApct.plot<-
  ggpredict(fitBApct,terms=c("Init [exp]","Species","BA.Target")) %>% 
  dplyr::rename(Species=group) %>% 
  ggplot(aes(x,predicted,color=Species))+
  geom_line()+
  facet_wrap(~facet)+
  ylab("Predicted 27-year basal area % increase")+
  xlab("Tree BA in 1992 (m²)")+
  theme(legend.position="bottom")

@@

# Raw plots
baAnn.raw.plot<-
  # Plot boxplot of increment by year
  piDat %>% 
  ggplot(aes(x=Year,y=Inc,color=BA.Target))+
  geom_boxplot()+
  facet_grid(Species~BA.Target)+
  xlab("Measurement year")+
  ylab("Annualized basal area increment (m²)")+
  theme(legend.position = "none")+
  theme(legend.position = "none")

@@

### Height-diameter models

```{r height diameter raw plot}

heightDiamRaw.plot<-
  dat %>% 
    filter(Species %in% c("Bl","Sx")) %>% # filter for Bl and Sx for now
    filter(Status=="Live") %>% # filter for live trees
    filter(is.na(DBT)) %>% # filter for trees with no broken tops
    drop_na(Height) %>% # remove trees where height was not measured

    # Plotting
  ggplot()+
  aes(x=DBH,y=Height,fill=Species)+
  geom_point()+
  facet_grid(Year~Species)+
  theme(legend.position = "none")+
  ylab("Height (m)")+
  xlab("DBH (cm)")+
  geom_smooth(formula = y~log(x),method="lm",se=FALSE)

# Fit model
htDiamFit<- 
  dat %>% 
    filter(Species %in% c("Bl","Sx")) %>% # filter for Bl and Sx for now
    filter(Status=="Live") %>% # filter for live trees
    filter(is.na(DBT)) %>% # filter for trees with no broken tops
    drop_na(Height) %>% # remove trees where height was not measured
  mutate(TreeID=paste(Plot,TreeID,sep="-")) %>% # create tree ID column
  lmer(Height~log(DBH)+(1|Plot/TreeID),data=.)

appNums("heightDiamRawPlot",display="FALSE")
```

Tree heights had a positive nonlinear relationship with diameter, for both species and across all time periods (`r appNums("heightDiamRawPlot",display="cite")`).  A comparison of height-diameter models showed that intercepts and slopes of log-transformed heights did not vary significantly (p<0.05) between species or time periods (not shown), therefore a single height-diameter linear mixed-effects model was developed using `r dat %>% filter(Species %in% c("Bl","Sx")) %>% filter(Status=="Live") %>% filter(is.na(DBT)) %>% drop_na(Height) %>% nrow()` tree height-diameter observations in the dataset, with log-transformed diameter as the fixed effect, and tree nested within plot as random effect. This model explained `r round(MuMIn::r.squaredGLMM(htDiamFit)[1]*100,1)`% of variation in log-transformed heights, with an intercept of `r round(fixef(htDiamFit)[1],3)` and slope coefficient of  `r round(fixef(htDiamFit)[2],3)` * log(DBH).  Residual plots were assessed to ensure goodness of fit.  This model was then applied to predict heights in trees without height measurements, and tree-volume estimates generated as per Nigh [-@Nigh2016].

@@

`r figNums("fitBA.Ann.plot",caption=paste("Predicted annualized basal area increment (m^2^ at breast height) over four periods from a linear mixed effects model explaining ", MuMIn::r.squaredGLMM(fitBA.Ann)[1] %>% round(3)*100, "% variation in log-transformed basal area increment with period, treatment unit and species as fixed effects (p<0.001 for all fixed effects, now shown). Basal area increment are average for growing seasons within each period.  For example, values at x-axis value of 1994 are mean for three growing seasons (between 1992 and 1994).  Predictions are back-transformed to original response scale.  Error bars represent 95% confidence intervals.",sep=""))`

@@

```{r plot height to diameter models by species and year, fig.width=6, fig.asp=1.2}

heightDiamRaw.plot

`r appNums("heightDiamRawPlot",caption="Height and diameter comparisons for 870 tree measurements, faceted by species and measurement year.")`
